---
title: "Montesinho Forest Fires"
author: "Fernando Anorve-Lopez and Christina Sousa"
date: "March 12, 2019"
output: html_document
references:
- id: cortezff
  title: A Data Mining Approach to Predict Forest Fires using Meteorological Data
  author:
  - family: Cortez
    given: Paulo
  - family: Morais
    given: Anibal
  container-title: New Trends in Artificial Intelligence, Proceedings of the 13th EPIA 2007 - Portuguese Conference on Artificial Intelligence
  editor:
    - family: Neves
      given: J
    - family: Santos
      given: M F
    - family: Machado
      given: J
  URL: 'http://www.dsi.uminho.pt/~pcortez/fires.pdf'
  publisher: Guimaraes, Portugal
  page: 512-523
  type: article-journal
  issued:
    year: 2017
    month: 12
---

#Introduction

The data are from forest fires occuring in the Montesinho National Park in northeast Portugal between January 2000 and December 2003. They come to us by way of Paulo Cortez and Anibal Morais at the University of Minho, in Guimaraes, Portugal. Our goal is to use 12 predictors to model the total burned area of large forest fires within the park. This could be done using a multiple linear regression model. Additionally, we would like to identify which values of the predictors yield fires with burn area less than $100m^2$, and which ones lead to larger fires. This could be done using logistic regression.

At each fire occurrence, a fire inspector recorded the date and and time of the fire, as well as the spatial location of the fire within the park boundaries. They also recorded the type of vegetation, weather conditions, total burn area, and six components of the Fire Weather Index (FWI), a Canadian system for rating fire danger. These included Fine Fuel Moisture Code (FFMC), Duff Moisture Code (DMC), Drought Code (DC), Initial Spread Index (ISI), Buildup Index (BUI) and FWI. The FFMC variable pertains to the moisture content of surface litter; the DMC and DC variable pertain to the "moisture content of shallow and deep organic layers" [cite main paper]; the ISI variable pertains to fire velocity spread; and finally the BUI variable pertains to the amount of available fuel. These various codes and scores each contribute to the final FWI score. Although the individual elements are not measured on the same scale, higher values indicate greater danger of fire in all cases. 

The researchers did not include BUI and FWI in the data, because these were strongly collinear with the other predictors.

The variables in the data [@cortezff] are given by:

   1. **X** *(nominal)*: x-axis spatial coordinate within the Montesinho park map: 1 to 9
   2. **Y** *(nominal)*: y-axis spatial coordinate within the Montesinho park map: 2 to 9
   3. **month** *(nominal)*: month of the year: "jan" to "dec" 
   4. **day** *(nominal)*: day of the week: "mon" to "sun"
   5. **FFMC** *(ordinal, continuous)*: FFMC index from the FWI system: 18.7 to 96.20
   6. **DMC** *(ordinal, continuous)*: DMC index from the FWI system: 1.1 to 291.3 
   7. **DC** *(ordinal, continuous)*: DC index from the FWI system: 7.9 to 860.6 
   8. **ISI** *(ordinal, continuous)*: ISI index from the FWI system: 0.0 to 56.10
   9. **temp** *(interval, continuous)*: temperature in Celsius degrees: 2.2 to 33.30
   10. **RH** *(ordinal, discrete)*: relative humidity in %: 15.0 to 100
   11. **wind** *(ratio, continuous)* wind speed in km/h: 0.40 to 9.40 
   12. **rain** *(ratio, discrete)*: outside rain in mm/m2 : 0.0 to 6.4 
   13. **area** *(ratio, continuous)*: the burned area of the forest (in ha): 0.00 to 1090.84 
  

    
    
#Exploratory Data Analysis

We begin the analysis by reading in the data, plotting it, and viewing relevant summary statistics.

```{r}
#data file can be obtained at https://archive.ics.uci.edu/ml/machine-learning-databases/forest-fires/
#read data into r
forest <-read.csv("forestfires.csv")

#change spatial variables X and Y to factors
forest$X<-as.factor(forest$X)
forest$Y<-as.factor(forest$Y)

#change RH variable to numeric rather than integer (don't think this is needed)
#forest$RH<-as.numeric(forest$RH)

head(forest)
pairs(forest)
```

Out of 517 records, there are 247 records with an `area` value of 0. The `area` data is considerably right-skewed. This is explained by Cortez and Morais as a general trend with forest fire data observed in various locations: either the fires are extinguished quickly and burn less than $100m^2.$ of area, or they are not easily contained and end up burning a large area. Fires that burn less than $100m^2$ of area are all catalogued as zero. This poses a missing data problem and prevents us from doing a complete analysis on these types of fires. To face this problem we can apply the MLR model to the data conditioned to nonzero burn area only. We remove these data for now, however, we will make use of them later to compare small fires and large fires using logistic regression.

```{r, echo = F}
hist(forest$area,main = "Wildfire Area",xlab = "Area")
forest2<-forest[which(forest$area>0),]
```


The values of `area` also range over several orders of magnitude.  The positive skewness suggests the need of a transformation. Trying log transformation seems suitable. We log transform `area` of the nonzero-valued observations and note that the histogram looks more symmetric now.

```{r, echo = F}
forest3<-cbind(forest2[,1:12],log(forest2$area))
colnames(forest3)[13]<-"Larea"
hist(forest3$Larea,main = "Wildfire Area",xlab = "Log(area)")
```

We now turn out attention to the other relevant summary statistics of the data. We notice that certain days and months appear to have more fires. Also, the distributions of `FFMC`, `DMC`, `DC`, and `ISI` appear to be somewhat skewed. Finally, `rain` does not seem to offer much information for forest fires with large areas.

```{r, echo=F}
summary(forest3)
```


As categorical data, the `month` variable can be divided into subsets defined by seasons. 

```{r, echo=F}
plot(forest3$month ,forest3$Larea, main = "Month vs. log(area)", ylab="log(area)")
``` 

We can also redefine Jan, Feb, Mar as Winter; Apr, May, Jun as Spring; Jul, Aug, Sep as Summer, and Oct, Nov, Dec, as Fall, because seasonal weather conditions can have an effect on incidence of wildfires. In this case, Summer records show more variability than any other season.

```{r, echo=F}

week <- rep(0,length(forest3$day))

week[which(forest3$day == "mon")] <- 0
week[which(forest3$day == "tue")] <- 0
week[which(forest3$day == "wed")] <- 0
week[which(forest3$day == "thu")] <- 0
week[which(forest3$day == "fri")] <- 0
week[which(forest3$day == "sat")] <- 1
week[which(forest3$day == "sun")] <- 1

seas <- rep(0,length(forest3$month), ylab = "log(area)",
     xlab = "Month",main = "Month Data")

seas[which(forest3$month == "jan")] <- 0
seas[which(forest3$month == "feb")] <- 0
seas[which(forest3$month == "mar")] <- 0
seas[which(forest3$month == "apr")] <- 1
seas[which(forest3$month == "may")] <- 1
seas[which(forest3$month == "jun")] <- 1
seas[which(forest3$month == "jul")] <- 2
seas[which(forest3$month == "aug")] <- 2
seas[which(forest3$month == "sep")] <- 2
seas[which(forest3$month == "oct")] <- 3
seas[which(forest3$month == "nov")] <- 3
seas[which(forest3$month == "dec")] <- 3

forest3$week <- as.factor(week)
forest3$season <- as.factor(seas)
```

```{r, echo=F}
plot(forest3$season ,forest3$Larea, ylab = "log(area)",
     xlab = "Season",main = "Seasonal Data",
     names = c("Winter","Spring","Summer","Winter"))
```

The `day` variable can be divided into subsets "weekdays" and "weekends", since forest fires could be explained by exposure to human beings who visit the park on the weekends. 

```{r, echo=F}
plot(forest3$day ,forest3$Larea, ylab = "log(area)",
     main = "Days")

plot(forest3$week ,forest3$Larea, ylab = "log(area)",
     main = "Days",
     names = c("Weekdays","Weekends"))
```


We also note that, after removing the forest fires with zero area, there are only 2 records with nonzero values for `rain`. Hence it does not make sense to include this predictor in our current analysis, and we remove it. We may take another look at this later in the logistic regression setting. We now view the continuous predictors after removal of this variable:

```{r, echo=F}
forest4<-forest3[,c(1:11,14,15,13)]
pairs(forest4[,c(5:11,14)], pch=18)

```

For the fire index variables, we note that the quantiles appear to be a bit skewed, particularly for `FFMC` and `DMC`, so we suspect that transformation of these variables may be in order. The boxplots of the fire index variables appear below.



```{r, echo=F}

par(mfrow=c(2,2))
for (i in c(5:8)){
  boxplot(forest4[,i], horizontal = TRUE, main = paste(colnames(forest4)[i]))
}
```

Noticing that `FFMC` is negatively skewed, we can try to normalize it by reflecting it, adding an appropriate value to make it positive, and finally applying a log transformation. $$f(\text{FFMC}) = \log(-\text{FFMC}+ \max(\text{FFMC})+1) $$

This seems to correct the skew. 

```{r, echo=F}
par(mfrow=c(1,1))
TrFFMC <- log(-forest4$FFMC+ max(forest4$FFMC)+1)
boxplot(TrFFMC, horizontal=T, main="Transformed FFMC")
plot(TrFFMC,forest4$Larea, xlab = "f(FFMC)" , ylab = "log(area)",
     main = "FFMC vs. log(Area)")
```


The final cleaned continuous variables appear below.

```{r, echo=F}
forest5<-forest4
forest5$FFMC<-TrFFMC
colnames(forest5)[5]<-"TrFFMC"
pairs(forest5[,c(5:11,14)], pch=18)
```

Finally, we comment on collinearity in this data. It appears that `DMC` and `DC` are linearly correlated. There may also be correlation between `temp` and `ISI` and `RH` and `temp`. In fact, `temp` appears to be linearly correlated with just about all the other predictors. There is also an interesting (and perhaps troubling!) curvilinear relationship between `DMC` and `ISI` as well as `DMC` and `temp`. To summarize, there appears quite a bit of multicollinearity in this data. This may be because the fire indexes are based on the various weather variables, and  because weather variables such as temperature and humidity are related to one another.

Summarizing our EDA, there do not appear to be any continuous variables that look like strong predictors for `Larea`. This leads us to think that a logistic regression approach might be more insightful to determine which variables are significant to predict if the burn area of a wildfire will be larger or smaller than $100m^2.$

To attempt to model `Larea`, we might proceed by seeking out suitable transformations of these variables by applying the Yeo-Johnson method. The `car` package function `powerTransform` suggests that the cleaned data could be normalized with a few reasonable transformations. Also, a regression tree suggests that `RH`, `DC`, `ISI`, and the transformed `FFMC` variables may still be significant in predicting `Larea`.


```{r, warning=F, message=F}
require(car)
summary(powerTransform(cbind(TrFFMC,DMC,DC,ISI,temp,RH,wind)~1,forest5,family="yjPower"))

```

```{r, message=F, warning=F}
require(tree)
data.tree <- tree(Larea~., data=forest5[,c(5:11,14)],mincut=15, minsize=35)
plot(data.tree, type="uniform")
text(data.tree)


```


#Citations
 